### 2020-07-11 paper.tex
checkpoint on probabilisic programming with traces

\section{INVOLUTIVE MCMC}

Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$.
Let $p : X \to [0, \infty)$ be a probability density with respect to $\mu_P$ and for each $t \in X$ such that $p(t) > 0$ let $q_t : Y \to [0, \infty)$ be probability density with respect to $\mu_Q$.
We call $p$ the \emph{target density} and we call $q_t$ \emph{auxiliary densities}.
Let $\pi(t, u) := p(t) q_t(u)$ and $Z := \{(t, u) : \pi(t, u) > 0 \}$.
Suppose $Z \in \Sigma_P \otimes \Sigma_Q$ (the product $\sigma$-algebra), so that $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and let $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$
(note that $\mu$ is $\sigma$-finite since $\mu_P \times \mu_Q$ is $\sigma$-finite. % TODO: prove this 
Suppose that $\pi$ is a probability density with respect to $\mu$. % TODO is this guaranteed?

Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
Let $u \sim q_t(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_t(u) \mu_Q(du)$ for $A \in \Sigma_Q$.

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $t$}
    \State $u \sim q_t(\cdot)$ \Comment{Sample auxiliary variables}
    \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
    %\State $\alpha \gets
        %\frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
    % TODO or should we show the Radon-Nikodym derivative?
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)$
    % TODO check it
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}
See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.

@@

\paragraph{Traces}
In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the name of a `random choice' to its value.
Following standard terminology in probabilistic programming~\citep{?}, we call these dictionaries \emph{traces} because they can be sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
In the probabilistic programming language used in this paper, the user assigns a unique name or \emph{address} to each random choice using a special syntax:

@@

TODO: introduce the mathematical notation for a trace

\paragraph{A probabilistic program encodes a probability distribution on traces}
For probabilistic programs that only sample individual random choices from discrete probability distributions, the meaning of the program can be described by a probability mass function that assigns a probailtiy to every possible trace.

@@

% measure version
\paragraph{Handling continuous distributions}
For programs that can sample from continuous probability distributions, the formalism needs to be extended.
pecifically, we define the $\mathcal{S}$ to be the set of all possible \emph{trace structures} as a (possibly countably infinite) set of finite sets of addresses, denoted $\mathcal{S} \subseteq \mathcal{P}(\mathcal{A})$, where $\mathcal{A}$ is the set of all addresses.
For each address $a \in \mathcal{A}$ we assign a reference measure space $(X_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite.
The set of all traces of a probabilistic program is the set $T := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$.
The \emph{reference measure space} for a probabilistic program is $(T, \Sigma, \mu)$ where
$\Sigma$ is the $\sigma$-algebra that contains sets $\cup_{S \in \mathcal{S}}\{ B_S \} \in \mathcal{P}(T)$ such that $B_S$ is an element of the product $\sigma$-algebra $\Sigma_S := \otimes_{a \in S} \Sigma_a$,
and where $\mu(\cup_{S \in \mathcal{S}}\{ B_S \}) := \sum_{S \in \mathcal{S}} \mu_S(B_S)$ where $\mu_S$ is the product measure $\times_{a \in S} \mu_a$.

We assume that a probabilistic program $P$ denotes a reference measure space $(T_P, \Sigma_P, \mu_P)$ and a measure $P(dt)$ that is mutually absolutely continuous with respect to $\mu_P$, with density 
$p : T \to [0, \infty)$ on traces $t \in T$.
We do not give the denotational semantics of a probabilistic programming language with this property, but instead refer to POPL paper for an example of a denotational semantics. % TODO what about requirement that the support of each random choice is constant? (this is implied by the definition of reference measure)

TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and the vbase measure and a density for it.

@@

\[
p(t) \;\; \pi(z)
\]

\paragraph{Extending one probabilistic program with another}
A second probabilistic program that is parametrized by (i.e. takes as input) traces of the first probabilistic program $P$.
The reference measure space of the second probabilistic program for $t \in T_P$ is denoted $(T_{Q,t}, \Sigma_{Q,t}, \mu_{P,t})$.
We define $Z := \{(t, u) : t \in T_P, u \in T_{Q,t}$ and we define $\pi(z) := p(t) q_t(u)$.
We require that $(Z, \Sigma, \mu)$ is a probabilistic program reference measure and $\pi$ is a probability density with respect to $\mu$ that is mutually absolutely continuous.
The combined program is $P \circ Q$.

Given a

% the only place where this matters is in the 'gibbs move' argument.. (8.4)

We need $p$ to be a density with respect to some reference measure space $(T_P, \Sigma_{P}, \mu_P)$.
We need $q_t$ to be a density with respect to some reference measure $\mu_{Q,t}$ for each $t \in T_P$.
Such that the function $\pi(t, u) := p(x) q_t(u)$ is a density $\pi(x, u)$ with respect to some reference measure $\mu$, and $\pi$ is mutually absolutely continuous with $\mu$.
(this can be constructed by using hierarchical addresses `model' and `proposal' for the random choics made in the two programs).

@@

\subsection{Proof of detailed balance for involution} \label{sec:involution-detailed-balance}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
We reproduce the result in our notation:
\begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}] \label{lemma:tierney-involution}
Let $(Z, \Sigma, \pi)$ denote a measure space.
Suppose $f$ is a one-to-one function from $Z$ onto $Z$ such that $f^{-1} = f$.
Consider the probability kernel $k$ defined by $k_z(A) := [f(z) \in A] \alpha(z, f(z)) + [z \in A] (1 - \alpha(z, f(z)))$ (where $\alpha(z, f(z))$ gives the probability of accepting a proposed transition from $z$ to $f(z)$).
Let $\nu(dz) := \pi(dz) + (\pi \circ f^{-1})(dz)$.
Let $h(z)$ be a density for $\pi$ with respect to $\nu$.
Let $A := \{z \in Z : h(z) > 0 \mbox{ and } h(f(z)) > 0\}$.
$k$ satisfies detailed balance with respect to $\pi$ if and only if:
\begin{enumerate}
\item $\alpha(z, f(z)) = 0$ for $\pi$-almost all $z \not \in A$
\item $\alpha(z, f(z)) \frac{h(z)}{h(f(z))} = \alpha(f(z), z)$
\end{enumerate}
\end{lemma}

>>>>

\section{INVOLUTIVE MCMC} \label{sec:involutive-mcmc}
Let $(X, \Sigma_P, \mu_P)$ and $(Y, \Sigma_Q, \mu_Q)$ denote two measure spaces with $\sigma$-finite $\mu_P$ and $\mu_Q$.
Let $p : X \to [0, \infty)$ be a probability density with respect to $\mu_P$ and for each $x \in X$ such that $p(x) > 0$ let $q_x : Y \to [0, \infty)$ be probability density with respect to $\mu_Q$.
We call $p$ the \emph{model density} and we call $q_x$ \emph{auxiliary densities}.
Let $\pi(x, y) := p(x) q_x(y)$ and $Z := \{(x, y) \in X \times Y : \pi(x, y) > 0 \}$.
Suppose that $Z \in \Sigma_P \otimes \Sigma_Q$ ($Z$ is a $\mu_P \times \mu_Q$-measurable set).
Then $\Sigma := \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq Z\}$ is a $\sigma$-algebra on $Z$ and $\mu : \Sigma \to [0, \infty)$ with $\mu(A) := (\mu_P \times \mu_Q)(A)$ is a $\sigma$-finite measure on measurable space $(Z, \Sigma)$.

Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.
Let $y \sim q_x(\cdot)$ denote a random sample from the measure defined by $A \mapsto \int_A q_x(y) \mu_Q(dy)$ for $A \in \Sigma_Q$.

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
    % TODO check it
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}
See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.

@@

\subsection{Traces}
In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the name of a `random choice' to its value.
Following standard terminology in probabilistic programming~\citep{?}, we call these dictionaries \emph{traces} because they can be sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
In the probabilistic programming language used in this paper, the user assigns a unique name or \emph{address} to each random choice using a special syntax:

@@

TODO: introduce the mathematical notation for a concrete trace (i.e. $\{a \mapsto 1, b \mapsto 2\}$).

\subsection{A probabilistic program encodes a probability distribution on traces}
For probabilistic programs that only sample individual random choices from discrete probability distributions, the meaning of the program can be described by a probability mass function that assigns a probailtiy to every possible trace.

@@

% measure version
\subsection{Handling Continuous Distributions}
For programs that can sample from continuous probability distributions, the formalism needs to be extended.
Let $\mathcal{A}$ denote a set of addresses.
For each address $a \in \mathcal{A}$ we assign a reference measure space $(V_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite ($V$ stands for `values').
Let $T := \cup_{A \subseteq \mathcal{A} : |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$ denote the set of all finite traces (i.e. dictionaries) from addresses in $\mathcal{A}$ to their values.
For each finite $A \subseteq \mathcal{A}$, let $\Sigma_A := \otimes_{a \in A} \Sigma_a$ (the product $\sigma$-algebra over $\Sigma_a$ for $a \in A$), and let $\mu_A := \times_{a \in A} \mu_a$ (the product measure).
Let $\Sigma \subseteq \mathcal{P}(T)$ be the $\sigma$-algebra formed by sets of the form $B := \cup_{A \subseteq \mathcal{A}, |A| < \infty} \{(A, \mathbf{x}) : \mathbf{x} \in B_A\}$ for some $B_A \in \Sigma_A$ for each $A$.
Let $\mu : \Sigma \to [0, \infty)$ be a measure on measurable space $(T, \Sigma)$ defined by $\mu(B) := \sum_{A \subseteq \mathcal{A}, |A| < \infty} \mu_A(B_A)$.

A probabilistic program $P$ denotes a probability density $p : T \to [0, \infty)$ with respect to a reference measure $\mu$ on traces.
Describing the denotational semantics of a particular probabilistic programming language (the function mapping programs $P$ to density functions $p$) is out of scope for this paper.
We refer to the reader to~\citet{popl} for an example of denotational semantics of a probabilistic programming language that uses explicit addresses for random-choices.

This formalism is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
For example, consider the following probabilistic program:

TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and show the reference measure on traces and its density.

@@

TODO: somewhere, talk about hierarchical addresses
%\paragraph{Hierarchical addresses}

\subsection{Involutive MCMC with probabilistic programs}

To define model the auxiliary densitites used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
The model reference measure $(X, \Sigma_P, \mu_P)$ and density required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure on traces as defined above and density $p$ (so that $X := T$).
To define the auxiliary densities, we require a probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
If both programs sample some random choice at the same address $a$ and require different measure spaces $(V_a, \Sigma_a, \mu_a)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}

Full support requirement (that for all structures and assignments to discrete random variables within the support, the other addresses have complete support).
Also for a pair of programs.
Use it guarantee that $X := \{(t, u) : \pi(t, u) > 0\} \in \Sigma_P \otimes \Sigma_Q$?

@@

\subsection{Proof of detailed balance for involution} \label{sec:involution-detailed-balance}

\begin{lemma}
Given $\sigma$-finite measures $(X_P, \Sigma_P, \mu_P)$ and $(X_Q, \Sigma_Q, \mu_Q)$,
if $X \in \Sigma_P \otimes \Sigma_Q$ then $\Sigma = \{A \in \Sigma_P \otimes \Sigma_Q : A \subseteq X\}$ is a $\sigma$-algebra on $X$, and $\mu : \Sigma \to [0, \infty)$ given by $\mu(A) := (\mu_P \times \mu_Q)(A)$ is $\sigma$-finite measure on measurable space $(X, \Sigma)$.
\end{lemma}
\begin{proof}
TODO: write out the detailed steps, proving each property of a $\sigma$-algebra%$ TODO
\end{proof}

\begin{lemma}
$\pi : X \to [0, \infty)$ is a probability density with respect to $\mu$ and the measure induced by $\pi$ and the measure $\mu$ are mutually absolutely continuous 
\end{lemma}
\begin{proof}
% TODO need to show that pi is measurable by the product measure?
\begin{equation}
\int_X \pi(x) \mu(dx) = \int_{X_P \times X_Q} \pi(x) (\mu_P \times \mu_Q)(dx) = \int_{X_P} p(t) \left( \int_{X_Q} q_t(u) \mu_Q(du) \right) \mu_P(dt) = 1
\end{equation}

%(note that $\mu$ is $\sigma$-finite since $\mu_P \times \mu_Q$ is $\sigma$-finite. % TODO: prove this 
%Suppose that $\pi$ is a probability density with respect to $\mu$. % TODO is this guaranteed?
\end{proof}

\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
We reproduce the result in our notation:
\begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}] \label{lemma:tierney-involution}
Let $(Z, \Sigma, \pi)$ denote a measure space.
Suppose $f$ is a one-to-one function from $Z$ onto $Z$ such that $f^{-1} = f$.
Consider the probability kernel $k$ defined by $k_z(A) := [f(z) \in A] \alpha(z, f(z)) + [z \in A] (1 - \alpha(z, f(z)))$ (where $\alpha(z, f(z))$ gives the probability of accepting a proposed transition from $z$ to $f(z)$).
Let $\nu(dz) := \pi(dz) + (\pi \circ f^{-1})(dz)$.
Let $h(z)$ be a density for $\pi$ with respect to $\nu$.
Let $A := \{z \in Z : h(z) > 0 \mbox{ and } h(f(z)) > 0\}$.
$k$ satisfies detailed balance with respect to $\pi$ if and only if:
\begin{enumerate}
\item $\alpha(z, f(z)) = 0$ for $\pi$-almost all $z \not \in A$
\item $\alpha(z, f(z)) \frac{h(z)}{h(f(z))} = \alpha(f(z), z)$
\end{enumerate}
\end{lemma}
